\textbf{Effective application of machine learning to microbiome-based classification problems}

### 2019-10-16 submission/manuscript.Rmd
sent to Pat for final review

Machine learning (ML) modeling of the human microbiome has the potential to identify microbial biomarkers and aid in the diagnosis of many diseases such as inflammatory bowel disease, diabetes, and colorectal cancer. Progress has been made towards developing ML models that predict health outcomes using bacterial abundances, but inconsistent adoption of training and evaluation methods call the validity of these models into question. Furthermore, there appears to be a preference by many researchers to favor increased model complexity over interpretability. To overcome these challenges, we trained seven models that used fecal 16S rRNA sequence data to predict the presence of colonic screen relevant neoplasias (SRNs; n=490 patients, 261 controls and 229 cases). We developed a reusable open-source pipeline to train, validate, and interpret the models. To show the effect of model selection, we assessed the predictive performance, interpretability, and training time of L2-regularized logistic regression, L1 and L2-regularized support vector machines (SVM) with linear and radial basis function kernels, decision trees, random forest, and gradient boosted trees (XGBoost). The random forest model performed best at detecting SRNs with an AUROC of 0.695 [IQR 0.651-0.739] but was slow to train (83.2 h) and not immediately interpretable. Despite its simplicity, L2-regularized logistic regression followed random forest in predictive performance with an AUROC of 0.680 [IQR 0.625-0.735], trained faster (12 min), and was inherently interpretable. Our analysis highlights the importance of choosing an ML approach based on the goal of the study, as the choice will inform expectations of performance and interpretability.

@@

To evaluate the generalizability of each model, we compared the median cross-validation AUROC to the median testing AUROC. If the difference between the cross-validation and testing AUROCs was large, then that could indicate that the models were overfit to the training data. The difference in median AUROCs was `r paste(round(abs(difference_model[difference_index[7],1]), 3))` in L1-regularized SVM with linear kernel, followed by SVM with radial basis function kernel and decision tree with a difference of `r paste(round(abs(difference_model[difference_index[6],1]), 3))` and `r paste(round(abs(difference_model[difference_index[5],1]), 3))`, respectively [Figure 2]. These differences are relatively small and give us confidence in our estimate of the generalization performance of the models.

@@

We established a pipeline that can be generalized to any modeling method that predicts a binary health outcome. We performed a random data-split to create a training set (80% of the data) and a held-out test set (20% of the data), which we used to evaluate predictive performance. We repeated this data-split 100 times to measure the possible variation in predictive performance. During training, we tuned the model hyperparameters with a repeated five-fold cross-validation. Despite the high number of features microbiome datasets typically have, the models we built with this pipeline generalized to the held-out test sets.

@@

**Data collection and study population.** The original stool samples described in our analysis were obtained from patients recruited by Great Lakes-New England Early Detection Research Network [@baxter_microbiota-based_2016]. Stool samples were provided by adults who were undergoing a scheduled screening or surveillance colonoscopy. Participants were recruited from Toronto (ON, Canada), Boston (MA, USA), Houston (TX, USA), and Ann Arbor (MI, USA). Patients’ colonic health was visually assessed by colonoscopy with bowel preparation and tissue histopathology of all resected lesions. We assigned patients into two classes: those with healthy colons and those with screen relevant neoplasias (SRNs). The healthy class included patients with healthy colons or non-advanced adenomas, whereas the SRN class included patients with advanced adenomas or carcinomas [@redwood_stool_2016]. Patients with an adenoma greater than 1 cm, more than three adenomas of any size, or an adenoma with villous histology were classified as having advanced adenomas. There were 172 patients with normal colonoscopies, 198 with adenomas, and 120 with carcinomas. Of the 198 adenomas, 109 were identified as advanced adenomas. Together 261 patients were classified as healthy and 229 patients were classified as having an SRN.

@@

**Statistical analysis workflow.** Data summaries, statistical analysis, and data visualizations were performed using R (v.3.5.0) with the tidyverse package (v.1.2.1). We compared the performance of the models pairwise by calculating the difference between AUROC values from the same datasplit (for 100 datasplits). We determined if the models were significantly different by calculating the empirical p-value  (2 x min(% of AUROC differences $\geq$ 0, % of AUROC differences $\leq$ 0) for the double tail event (e.g., Figure S3).

@@

**Figure 3. Interpretation of the linear ML models.** The ranks of absolute feature weights of (A) L1-regularized SVM with linear kernel, (B) L2-regularized SVM with linear kernel, and (C) L2-regularized logistic regression, were ranked from highest rank, 1, to lowest rank, 100, for each datasplit. The feature ranks of the 20 highest ranked OTUs based on their median ranks (median shown in black) are reported here.  OTUs that were associated with classifying a subject as being healthy had negative signs and were shown in blue. OTUs that were associated with classifying a subject having an SRN had positive signs and were shown in red. Some of the same OTUs were identified as important in all of the linear models. 
\newpage
\includegraphics{Figure_4.png}

@@

**Figure S3. Histogram of AUROC differences between L2-regularized logistic regression and random forest for each of the hundred datasplits.** This histogram shows the number of datasplits in each bin. The percentage of dataplits where the difference between random forest and L2-regularized logistic regression AUROC values was higher than or equal to 0 were 0.75, lower than or equal to 0 were 0.25. The vertical red line highlights the bins where there AUROC difference between the two model is 0. The p-value was calculated for a double tail event. 
\newpage
\includegraphics[height=17.5cm, width=13cm]{Figure_S4.png}

>>>>

Machine learning (ML) modeling of the human microbiome has the potential to identify microbial biomarkers and aid in the diagnosis of many diseases such as inflammatory bowel disease, diabetes, and colorectal cancer. Progress has been made towards developing ML models that predict health outcomes using bacterial abundances, but inconsistent adoption of training and evaluation methods call the validity of these models into question. Furthermore, there appears to be a preference by many researchers to favor increased model complexity over interpretability. To overcome these challenges, we trained seven models that used fecal 16S rRNA sequence data to predict the presence of colonic screen relevant neoplasias (SRNs; n=490 patients, 261 controls and 229 cases). We developed a reusable open-source pipeline to train, validate, and interpret ML models. To show the effect of model selection, we assessed the predictive performance, interpretability, and training time of L2-regularized logistic regression, L1 and L2-regularized support vector machines (SVM) with linear and radial basis function kernels, decision trees, random forest, and gradient boosted trees (XGBoost). The random forest model performed best at detecting SRNs with an AUROC of 0.695 [IQR 0.651-0.739] but was slow to train (83.2 h) and not immediately interpretable. Despite its simplicity, L2-regularized logistic regression followed random forest in predictive performance with an AUROC of 0.680 [IQR 0.625-0.735], trained faster (12 min), and was inherently interpretable. Our analysis highlights the importance of choosing an ML approach based on the goal of the study, as the choice will inform expectations of performance and interpretability.

@@

To evaluate the generalizability of each model, we compared the median cross-validation AUROC to the median testing AUROC. If the difference between the cross-validation and testing AUROCs was large, then that could indicate that the models were overfit to the training data. The largest difference in median AUROCs was `r paste(round(abs(difference_model[difference_index[7],1]), 3))` in L1-regularized SVM with linear kernel, followed by SVM with radial basis function kernel and decision tree with a difference of `r paste(round(abs(difference_model[difference_index[6],1]), 3))` and `r paste(round(abs(difference_model[difference_index[5],1]), 3))`, respectively [Figure 2]. These differences are relatively small and give us confidence in our estimate of the generalization performance of the models.

@@

We established a pipeline that can be generalized to any modeling method that predicts a binary health outcome. We performed a random data-split to create a training set (80% of the data) and a held-out test set (20% of the data), which we used to evaluate predictive performance. We used the AUROC metric to evaluate predictive performance as our dataset was highly balanced. We repeated this data-split 100 times to measure the possible variation in predictive performance. During training, we tuned the model hyperparameters with a repeated five-fold cross-validation. Despite the high number of features microbiome datasets typically have, the models we built with this pipeline generalized to the held-out test sets.

@@

**Data collection and study population.** The original stool samples described in our analysis were obtained from patients recruited by Great Lakes-New England Early Detection Research Network [@baxter_microbiota-based_2016]. Stool samples were provided by adults who were undergoing a scheduled screening or surveillance colonoscopy. Participants were recruited from Toronto (ON, Canada), Boston (MA, USA), Houston (TX, USA), and Ann Arbor (MI, USA). Patients’ colonic health was visually assessed by colonoscopy with bowel preparation and tissue histopathology of all resected lesions. We assigned patients into two classes: those with healthy colons and those with screen relevant neoplasias (SRNs). The healthy class included patients with healthy colons or non-advanced adenomas, whereas the SRN class included patients with advanced adenomas or carcinomas [@redwood_stool_2016]. Patients with an adenoma greater than 1 cm, more than three adenomas of any size, or an adenoma with villous histology were classified as having advanced adenomas [@redwood_stool_2016]. There were 172 patients with normal colonoscopies, 198 with adenomas, and 120 with carcinomas. Of the 198 adenomas, 109 were identified as advanced adenomas. Together 261 patients were classified as healthy and 229 patients were classified as having an SRN.

@@

**Statistical analysis workflow.** Data summaries, statistical analysis, and data visualizations were performed using R (v.3.5.0) with the tidyverse package (v.1.2.1). We compared the performance of the models pairwise by calculating the difference between AUROC values from the same data-split (for 100 data-splits). We determined if the models were significantly different by calculating the empirical p-value  (2 x min(% of AUROC differences $\geq$ 0, % of AUROC differences $\leq$ 0) for the double tail event (e.g., Figure S3).

@@

**Figure 3. Interpretation of the linear ML models.** The ranks of absolute feature weights of (A) L1-regularized SVM with linear kernel, (B) L2-regularized SVM with linear kernel, and (C) L2-regularized logistic regression, were ranked from highest rank, 1, to lowest rank, 100, for each data-split. The feature ranks of the 20 highest ranked OTUs based on their median ranks (median shown in black) are reported here.  OTUs that were associated with classifying a subject as being healthy had negative signs and were shown in blue. OTUs that were associated with classifying a subject having an SRN had positive signs and were shown in red. Some of the same OTUs were identified as important in all of the linear models. 
\newpage
\includegraphics{Figure_4.png}

@@

**Figure S3. Histogram of AUROC differences between L2-regularized logistic regression and random forest for each of the hundred data-splits.** This histogram shows the number of data-splits in each bin. The percentage of dataplits where the difference between random forest and L2-regularized logistic regression AUROC values was higher than or equal to 0 were 0.75, lower than or equal to 0 were 0.25. The vertical red line highlights the bins where there AUROC difference between the two model is 0. The p-value was calculated for a double tail event. 
\newpage
\includegraphics[height=17.5cm, width=13cm]{Figure_S4.png}

### 2019-10-16 submission/manuscript.tex
sent to Pat for final review

Machine learning (ML) modeling of the human microbiome has the potential
to identify microbial biomarkers and aid in the diagnosis of many
diseases such as inflammatory bowel disease, diabetes, and colorectal
cancer. Progress has been made towards developing ML models that predict
health outcomes using bacterial abundances, but inconsistent adoption of
training and evaluation methods call the validity of these models into
question. Furthermore, there appears to be a preference by many
researchers to favor increased model complexity over interpretability.
To overcome these challenges, we trained seven models that used fecal
16S rRNA sequence data to predict the presence of colonic screen
relevant neoplasias (SRNs; n=490 patients, 261 controls and 229 cases).
We developed a reusable open-source pipeline to train, validate, and
interpret the models. To show the effect of model selection, we assessed
the predictive performance, interpretability, and training time of
L2-regularized logistic regression, L1 and L2-regularized support vector
machines (SVM) with linear and radial basis function kernels, decision
trees, random forest, and gradient boosted trees (XGBoost). The random
forest model performed best at detecting SRNs with an AUROC of 0.695
{[}IQR 0.651-0.739{]} but was slow to train (83.2 h) and not immediately
interpretable. Despite its simplicity, L2-regularized logistic
regression followed random forest in predictive performance with an
AUROC of 0.680 {[}IQR 0.625-0.735{]}, trained faster (12 min), and was
inherently interpretable. Our analysis highlights the importance of
choosing an ML approach based on the goal of the study, as the choice
will inform expectations of performance and interpretability.

@@

To evaluate the generalizability of each model, we compared the median
cross-validation AUROC to the median testing AUROC. If the difference
between the cross-validation and testing AUROCs was large, then that
could indicate that the models were overfit to the training data. The
difference in median AUROCs was 0.021 in L1-regularized SVM with linear
kernel, followed by SVM with radial basis function kernel and decision
tree with a difference of 0.007 and 0.006, respectively {[}Figure 2{]}.
These differences are relatively small and give us confidence in our
estimate of the generalization performance of the models.

@@

We established a pipeline that can be generalized to any modeling method
that predicts a binary health outcome. We performed a random data-split
to create a training set (80\% of the data) and a held-out test set
(20\% of the data), which we used to evaluate predictive performance. We
repeated this data-split 100 times to measure the possible variation in
predictive performance. During training, we tuned the model
hyperparameters with a repeated five-fold cross-validation. Despite the
high number of features microbiome datasets typically have, the models
we built with this pipeline generalized to the held-out test sets.

@@

\textbf{Data collection and study population.} The original stool
samples described in our analysis were obtained from patients recruited
by Great Lakes-New England Early Detection Research Network (5). Stool
samples were provided by adults who were undergoing a scheduled
screening or surveillance colonoscopy. Participants were recruited from
Toronto (ON, Canada), Boston (MA, USA), Houston (TX, USA), and Ann Arbor
(MI, USA). Patients' colonic health was visually assessed by colonoscopy
with bowel preparation and tissue histopathology of all resected
lesions. We assigned patients into two classes: those with healthy
colons and those with screen relevant neoplasias (SRNs). The healthy
class included patients with healthy colons or non-advanced adenomas,
whereas the SRN class included patients with advanced adenomas or
carcinomas (43). Patients with an adenoma greater than 1 cm, more than
three adenomas of any size, or an adenoma with villous histology were
classified as having advanced adenomas. There were 172 patients with
normal colonoscopies, 198 with adenomas, and 120 with carcinomas. Of the
198 adenomas, 109 were identified as advanced adenomas. Together 261
patients were classified as healthy and 229 patients were classified as
having an SRN.

@@

\textbf{Statistical analysis workflow.} Data summaries, statistical
analysis, and data visualizations were performed using R (v.3.5.0) with
the tidyverse package (v.1.2.1). We compared the performance of the
models pairwise by calculating the difference between AUROC values from
the same datasplit (for 100 datasplits). We determined if the models
were significantly different by calculating the empirical p-value (2 x
min(\% of AUROC differences \(\geq\) 0, \% of AUROC differences \(\leq\)
0) for the double tail event (e.g., Figure S3).

@@

\textbf{Figure 3. Interpretation of the linear ML models.} The ranks of
absolute feature weights of (A) L1-regularized SVM with linear kernel,
(B) L2-regularized SVM with linear kernel, and (C) L2-regularized
logistic regression, were ranked from highest rank, 1, to lowest rank,
100, for each datasplit. The feature ranks of the 20 highest ranked OTUs
based on their median ranks (median shown in black) are reported here.
OTUs that were associated with classifying a subject as being healthy
had negative signs and were shown in blue. OTUs that were associated
with classifying a subject having an SRN had positive signs and were
shown in red. Some of the same OTUs were identified as important in all
of the linear models. \newpage
\includegraphics{Figure_4.png}

@@

\textbf{Figure S3. Histogram of AUROC differences between L2-regularized
logistic regression and random forest for each of the hundred
datasplits.} This histogram shows the number of datasplits in each bin.
The percentage of dataplits where the difference between random forest
and L2-regularized logistic regression AUROC values was higher than or
equal to 0 were 0.75, lower than or equal to 0 were 0.25. The vertical
red line highlights the bins where there AUROC difference between the
two model is 0. The p-value was calculated for a double tail event.
\newpage
\includegraphics[height=17.5cm, width=13cm]{Figure_S4.png}

>>>>

Machine learning (ML) modeling of the human microbiome has the potential
to identify microbial biomarkers and aid in the diagnosis of many
diseases such as inflammatory bowel disease, diabetes, and colorectal
cancer. Progress has been made towards developing ML models that predict
health outcomes using bacterial abundances, but inconsistent adoption of
training and evaluation methods call the validity of these models into
question. Furthermore, there appears to be a preference by many
researchers to favor increased model complexity over interpretability.
To overcome these challenges, we trained seven models that used fecal
16S rRNA sequence data to predict the presence of colonic screen
relevant neoplasias (SRNs; n=490 patients, 261 controls and 229 cases).
We developed a reusable open-source pipeline to train, validate, and
interpret ML models. To show the effect of model selection, we assessed
the predictive performance, interpretability, and training time of
L2-regularized logistic regression, L1 and L2-regularized support vector
machines (SVM) with linear and radial basis function kernels, decision
trees, random forest, and gradient boosted trees (XGBoost). The random
forest model performed best at detecting SRNs with an AUROC of 0.695
{[}IQR 0.651-0.739{]} but was slow to train (83.2 h) and not immediately
interpretable. Despite its simplicity, L2-regularized logistic
regression followed random forest in predictive performance with an
AUROC of 0.680 {[}IQR 0.625-0.735{]}, trained faster (12 min), and was
inherently interpretable. Our analysis highlights the importance of
choosing an ML approach based on the goal of the study, as the choice
will inform expectations of performance and interpretability.

@@

To evaluate the generalizability of each model, we compared the median
cross-validation AUROC to the median testing AUROC. If the difference
between the cross-validation and testing AUROCs was large, then that
could indicate that the models were overfit to the training data. The
largest difference in median AUROCs was 0.021 in L1-regularized SVM with
linear kernel, followed by SVM with radial basis function kernel and
decision tree with a difference of 0.007 and 0.006, respectively
{[}Figure 2{]}. These differences are relatively small and give us
confidence in our estimate of the generalization performance of the
models.

@@

We established a pipeline that can be generalized to any modeling method
that predicts a binary health outcome. We performed a random data-split
to create a training set (80\% of the data) and a held-out test set
(20\% of the data), which we used to evaluate predictive performance. We
used the AUROC metric to evaluate predictive performance as our dataset
was highly balanced. We repeated this data-split 100 times to measure
the possible variation in predictive performance. During training, we
tuned the model hyperparameters with a repeated five-fold
cross-validation. Despite the high number of features microbiome
datasets typically have, the models we built with this pipeline
generalized to the held-out test sets.

@@

\textbf{Data collection and study population.} The original stool
samples described in our analysis were obtained from patients recruited
by Great Lakes-New England Early Detection Research Network (5). Stool
samples were provided by adults who were undergoing a scheduled
screening or surveillance colonoscopy. Participants were recruited from
Toronto (ON, Canada), Boston (MA, USA), Houston (TX, USA), and Ann Arbor
(MI, USA). Patients' colonic health was visually assessed by colonoscopy
with bowel preparation and tissue histopathology of all resected
lesions. We assigned patients into two classes: those with healthy
colons and those with screen relevant neoplasias (SRNs). The healthy
class included patients with healthy colons or non-advanced adenomas,
whereas the SRN class included patients with advanced adenomas or
carcinomas (43). Patients with an adenoma greater than 1 cm, more than
three adenomas of any size, or an adenoma with villous histology were
classified as having advanced adenomas (43). There were 172 patients
with normal colonoscopies, 198 with adenomas, and 120 with carcinomas.
Of the 198 adenomas, 109 were identified as advanced adenomas. Together
261 patients were classified as healthy and 229 patients were classified
as having an SRN.

@@

\textbf{Statistical analysis workflow.} Data summaries, statistical
analysis, and data visualizations were performed using R (v.3.5.0) with
the tidyverse package (v.1.2.1). We compared the performance of the
models pairwise by calculating the difference between AUROC values from
the same data-split (for 100 data-splits). We determined if the models
were significantly different by calculating the empirical p-value (2 x
min(\% of AUROC differences \(\geq\) 0, \% of AUROC differences \(\leq\)
0) for the double tail event (e.g., Figure S3).

@@

\textbf{Figure 3. Interpretation of the linear ML models.} The ranks of
absolute feature weights of (A) L1-regularized SVM with linear kernel,
(B) L2-regularized SVM with linear kernel, and (C) L2-regularized
logistic regression, were ranked from highest rank, 1, to lowest rank,
100, for each data-split. The feature ranks of the 20 highest ranked
OTUs based on their median ranks (median shown in black) are reported
here. OTUs that were associated with classifying a subject as being
healthy had negative signs and were shown in blue. OTUs that were
associated with classifying a subject having an SRN had positive signs
and were shown in red. Some of the same OTUs were identified as
important in all of the linear models. \newpage
\includegraphics{Figure_4.png}

@@

\textbf{Figure S3. Histogram of AUROC differences between L2-regularized
logistic regression and random forest for each of the hundred
data-splits.} This histogram shows the number of data-splits in each
bin. The percentage of dataplits where the difference between random
forest and L2-regularized logistic regression AUROC values was higher
than or equal to 0 were 0.75, lower than or equal to 0 were 0.25. The
vertical red line highlights the bins where there AUROC difference
between the two model is 0. The p-value was calculated for a double tail
event. \newpage
\includegraphics[height=17.5cm, width=13cm]{Figure_S4.png}
